Containers and R---the Rockerverse and beyond

### 2020-01-16 manuscript.tex
update manuscript

% !TeX root = RJwrapper.tex
\title{Containers and R---the Rockerverse and beyond}
\author{by Daniel Nüst, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Karthik Ram, Noam Ross, Nan Xiao, Lori Shepherd, Nitesh Turaga, Jacqueline Nolis, Heather Nolis, Hong Ooi, Ellis Hughes, Sean Lopp, Ben Marwick}

@@

\abstract{%
The Rocker project provides widely-used Docker images for R across
different application scenarios. This articles surveys downstream
projects building upon Rocker and presents the current state of R
packages for managing Docker images and controlling containers. We also
look beyond Rocker to other projects connecting containerisation with R,
namely alternative suites of images. These use cases and the variety of
applications demonstrate the power of Rocker specifically and
containerisation in general. We identified common themes across this
diversity: reproducible environments, scalability and effectivity, and
portability across clouds.
}

@@

The R community keeps growing. This can be seen in the number of new
packages on CRAN, which keeps on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies taking up,
as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. Coinciding
with the rise of R was the advent of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distribution and deployment of server applications---in
fact, Docker can be called the \emph{lingua~franca} of describing
computing environments and packaging software. Combining both these
topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides images with R (see the
next Section for more details). The considerable uptake and continued
evolution of the Rocker Project has lead to numerous projects extending
or building upon Rocker images, ranging from reproducible research to
production deployments. This article presents this \emph{Rockerverse} of
projects across all development stages: early demonstrations, working
prototypes, and mature products. We also introduce related activities
connecting the R language and environment with other containerisation
solutions. The main contributions is a coherent picture of the current
lay of the land of using containers in, with, and for R.

The article continues with a brief introduction of containerization
basics and the Rocker Project, followed by descriptions of projects
developing images with R installations. Then, we present applications,
starting with the R packages specificly for interacting with Docker,
second-level packages using containers mediately or only for specific
features, up to complex use cases leveraging containers. We conclude
with a reflection on the landscape of packages and applications and
point out future directions of development.

@@

Docker, an application and service provide by the eponymous company, has
in just a few short years risen to prominence for development, testing,
deployment and distribution of computer software
\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While there are
related approaches such as
LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
or Singularity \citep{kurtzer_singularity_2017}, Docker has become
synomymous with ``containerization''---the method of taking software
artefacts and bundling them in such a way that use becomes standardized
and portable across operationg systems. In doing so, Docker had
recognised and validated the importance of one very important thread
that had been emerging, namely virtualization. By allowing (one or
possibly) multiple applications or services to run concurrently on one
host machine without any fear of interference between them, an important
scalability opportunity is being provided. But Docker improved this
compartmentalization by accessing the host system---generally
Linux---through a much thinner and smaller shim than a full operating
system emulation or virtualization. This containerization is also called
operating-system-level virtualization
\ref{wikipedia_contributors_os-level_2020}. Typically a container runs
one process, whereas virtualization may run whole operating systems at a
larger footprint. This makes for more efficient use of system resources
\citep{felter_updated_2015} and allowed another order of magnitude in
terms of scalability of deployment \citep[cf.][]{datadog_8_2018}. While
Docker makes use of Linux kernel features, its importance was large
enough so that some required aspects of running Docker have been added
to other operating systems to support Docker there more efficiently too
\citep{microsoft_linux_2019}. The success even lead to standardisation
and industry collaboration \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the Rocker
Project has provided a variety of Docker containers since its start in
2014 \citep{RJ-2017-065}. The Rocker Project provides several lines of
containers spanning to from building blocks with \texttt{R-release} or
\texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for better maintainability
(i.e.~smaller \texttt{Dockerfile}s), composability, and to reduce build
time. Also of note is a series of ``versioned'' containers which match
the R release they contain with the \emph{then-current} set of packages
via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, who provide findung for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled
variety of established and upcoming features. Nevertheless, special use
cases and experimental projects exist to test approaches or provide
features different to what \emph{``base R''} provides. These projects
stem both form academia and industry. \emph{Base R}, sometimes called
\texttt{GNU-R}, is the R distribution maintained by the
\href{https://www.r-project.org/contributors.html}{R Core Team} and and
provided via \href{https://cran.r-project.org/}{CRAN}.

\textbf{Microsoft R Open}
(\href{https://github.com/nuest/mro-docker}{MRO}) is an R distribution
formerly known as Revolution R Open (RRO) before Revolution Analytics
was acquired by Microsoft. MRO is compatible with main R and its
packages. ``It includes additional capabilities for improved
performance, reproducibility, and platform support.''
\citep{microsoft_mro} Most notably these capabilities are the MRAN
repository, which is enabled by default, and the (optional) integration
with \href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel
Library} (MKL) for multi-threading in linear algebra operations
\citep{microsoft_multithread}. MRO does not provide official Docker
images, but a set of community-maintained \texttt{Dockerfile}s and
Docker images \texttt{nuest/mro} are available on Docker Hub at
\url{https://github.com/nuest/mro-docker} and on GitHub at
\url{https://github.com/nuest/mro-docker} respectively. The images are
inspired by the Rocker images and can be used much in the same fashion,
effectively a drop-in replacement allowing users to quickly evaluate if
the benefits of MRO + Intel® MKL apply to their use case. Version-tagged
images are provided for the latest bugfix release of recent R versions.
Extended license information about MKL is printed at every startup.

\textbf{Renjin} is an interpreter for the R language running on top of
the \href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java
Virtual Machine} (JVM) providing full two-way access between Java and R
code \citep{wikipedia_renjin_2018}. It was developed to combine the
benefits of R, such scripting and extension packages, with the JVM's
advantages in the areas of security, cross-platform availability, and
established position in enterprise settings. R extension packages need
to be specially compiled and are distributed via the Java package
manager \href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
cf.~\url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e., at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g., \texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained
images for selected releases only are available under
\texttt{nuest/renjin} on Docker Hub and GitHub at
\url{https://hub.docker.com/r/nuest/renjin} and
\url{https://github.com/nuest/renjin-docker} respectively. These images
expose the command line interface of Renjin in a similar fashion as
Rocker images and allow an easy evaluation of Renjin's suitability, but
are not intended for production use.

\textbf{\href{http://www.pqr-project.org/}{pqR}} is a ``a pretty quick
version of R''. \emph{pqR} attempts to improve R on some opinionated
issues in the R language and is the basis for experimental features,
e.g., automatic
diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development
\href{https://github.com/radfordneal/pqR/}{on GitHub} is a one man
project and it does not provide any Docker images. But especially
disruptive approaches may contribute to the development of the R
ecosystem, so the \texttt{nuest/pqr} images on Docker Hub and GitHub at
\url{https://hub.docker.com/r/nuest/pqr/} and
\url{https://github.com/nuest/pqr-docker} respectively.

\textbf{FastR} is ``A high-performance implementation of the R
programming language, built on GraalVM''
(\url{https://github.com/oracle/fastr}). It is developed by Oracle,
connects R to the GraalVM ecosystem \citep{wikipedia_graalvm_2019}, and
also claims superior performance but also targets full compatibility
with base
R\footnote{\href{https://github.com/oracle/fastr}{https://github.com/oracle/fastr}}.
There are no offical Docker images provided, but \texttt{nuest/fastr}
images and \texttt{Dockerfile} are independently maintained on
\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{Docker
Hub} and \href{https://github.com/nuest/fastr-docker}{GitHub}
respectively.

While the images presented in this section are far from being as vetted,
stable, and widely used as any of the Rocker images, they demonstrate an
important advantage of containerisation technology, namely the abiliy to
transparently build portable stacks of open source software and make
them easily accessible to users. All different distributions are
published under GPL licenses. Since all of the different R distributions
claim better performance as a core motivation, a comparision based on
Docker images, potentially leveraging the
\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
restriction mechanisms} of Docker to level the playing field, seems
useful future work.

@@

\emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
open development project for the analysis and comprehension of genomic
data \citep{gentleman_bioconductor_2004}. The project consists of 1741 R
software packages as of August 15th 2019, as well as packages containing
annotation or experiment data. \_Bioconductor\_has a semiannual release
cycle, each release is associated with a particular version of R. Docker
images allow availability of current and past versions of
\emph{Bioconductor} for convenience and reproducibility.
\emph{Bioconductor} `base' docker images are built on top of
\texttt{rocker/r-ver} and \texttt{rocker/rstudio}. \emph{Bioconductor}
installs packages based on the R version, and therefore uses
\texttt{rocker/rstudio} and \texttt{rocker/r-ver} version tagging.
\emph{Bioconductor} selects the desired version of R from Rocker, adds
the BiocManager CRAN package for installing appropriate versions of
\emph{Bioconductor} packages, and creates a \emph{Bioconductor} docker
image with an informative tag (R\_version\_Bioc\_version). The images
are summarized on the \emph{Bioconductor} web site
(\url{https://bioconductor.org/help/docker/}), maintained on GitHub
(\url{https://github.com/Bioconductor/bioc_docker}), and available to
the community through \href{https://hub.docker.com/}{DockerHub}. Past
and current combinations of R and \emph{Bioconductor} are therefore
accessible via a specific docker tag.

@@

Use of images suggests several interesting possibilities for the
\emph{Bioconductor} project. Images may be valuable in teaching, where
participants pull pre-built images to avoid complicated configuration of
their own computing environemnts. An appeal of this over our current
approach (providing Amazon Machine Instances for the duration of the
course) is the utility of the image to participant after the course is
over. \texttt{bioconductor\_full} introduces a common system
configuration, so it becomes increasingly sensible for
\emph{Bioconductor} to distribute convenient \emph{binary} packages.
Images also suggest approaches to more advanced computational models.
For instance, we are exploring use of images for
\href{https://helm.sh/}{Helm}-orchestrated
\href{https://kubernetes.io/}{Kubernetes} clusters on the Google Cloud
Platform. The user interacts with a manager image based on
\texttt{bioconductor\_full}, configured to perform map-reduce style
computations via the BiocParallel package communicating with
minimally-configured worker images. A strength of this approach is that
the responsibility for complex software configuration (including
customized development) is shifted from the user to the experienced
\emph{Bioconductor} core team.

@@

Docker containers on the Windows operating system were originally quite
cumbersome to use. They required an extra tool,
\href{https://docs.docker.com/toolbox/}{Docker Toolbox}, which by now
only exists as a legacy solution for older Windows systems. Docker
toolbox leverages \texttt{docker-machine} (see also
Section~\nameref{interfaces} for an R package interfacing with
\texttt{docker-machine}) to handle the process of creating a local
virtual host which could host Docker Engine, while exposing the regular
Docker CLI. The Docker CLI commands are forwarded to the virtual host
transparently for the user. For current Windows Server (2016 and later)
and Windows Desktop
(\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
requires Windows 10) versions, Docker is supported
natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}}
and different base images are offered by
Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
The Docker CLI can be used in just the same way as on other operating
systems but not every base image is supported on every Windows
host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
On Docker Desktop for Windows, the user can run both Linux-based and
Windows-based containers, but only one of the Docker daemons can be used
at a
time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.

\textbf{rocker-win} (\url{https://github.com/nuest/rocker-win}) is a
proof of concept for running R in Windows-based containers
\citep{nust_rocker-win_2019}. It provides selected R versions using
three different base images using \texttt{microsoft/windowsservercore}
and \texttt{mcr.microsoft.com/windows/servercore}. These base images
match different Windows versions: Windows Server 2019, Windows Server
2016, and Windows Server, version 1803. The images are built using two
CI services,
\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
which provide the different Windows versions and therefore support
different base images, and published automatically
\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker Hub}. The
images are built from manually maintained seperate \texttt{Dockerfile}s,
are tagged with both Windows-variant and R version, e.g.,
\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
process by default. New R versions are only added on demand. The
explorative project demonstrates the variety and pontentially confusing
complexity of Windows on Docker, e.g., because of the different base
images, but also provides a starting point for any Windows-constrained
user to leverage containerisation, including examples for apps using
\textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
dependencies such as \pkg{sf}. The latter is quite comfortable actually,
as R ships pre-compiled binaries for Windows.

For Windows, just as originally for Linux, the cloud use cases drive the
development and Docker and Microsoft collaborate
closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about
reproducibility, will prefer the more widely and host independent
Linux-based containers. Naturally, a Windows license is required for the
host machine. But the developments in the \texttt{rocker-win} prototype
show that containers can also be used for R workflows depending on
Windows-only tools, for leveraging an existing Windows Server-based
infrastructure where policies can otherwise not be met, or for
streamlining interactions with system operating staff that runs
Windows-based containers themselves.

@@

\begin{itemize}
\tightlist
\item
  \href{https://rstudio.github.io/renv/articles/docker.html}{\texttt{renv}
  with Docker}
\end{itemize}

@@

  \begin{itemize}
  \tightlist
  \item
    For this project, we use travis-ci to build rocker-derived
    containers, test them, and only push them to docker hub (from
    travis-ci.org) if the integration tests succeed.
  \end{itemize}
\item
  Google Cloud Build

@@

\textbf{\href{https://github.com/att/rcloud/tree/master/docker}{RCloud}}
uses a \texttt{rocker/drd} base image for creation of collaborative data
analysis and visualisation environments.

\hypertarget{using-docker-in-a-commercial-data-science-platform}{%
\subsection{Using Docker in a commercial Data Science
Platform}\label{using-docker-in-a-commercial-data-science-platform}}

@@

\hypertarget{processing-div}{%
\subsection{Processing {[}div{]}}\label{processing-div}}

>>>>

% !TeX root = RJwrapper.tex
\title{Containers and R---the Rockerverse and beyond}
\author{by Daniel Nüst, Dom Bennett, Robrecht Cannoodt, Dirk Eddelbuettel, Mark Edmondson, Colin Fay, Ellis Hughes, Sean Lopp, Ben Marwick, Heather Nolis, Jacqueline Nolis, Hong Ooi, Karthik Ram, Noam Ross, Lori Shepherd, Nitesh Turaga, Nan Xiao}

@@

\abstract{%
The Rocker~Project provides widely-used Docker images for R across
different application scenarios. This articles surveys downstream
projects building upon Rocker and presents the current state of R
packages for managing Docker images and controlling containers. We also
look beyond Rocker to other projects connecting containerisation with R,
namely alternative suites of images. These use cases and the variety of
applications demonstrate the power of Rocker specifically and
containerisation in general. We identified common themes across this
diversity: reproducible environments, scalability and effectivity, and
portability across clouds.
}

@@

The R community keeps growing. This can be seen in the number of new
packages on CRAN, which keeps on growing exponentially
\citep{cran:2019}, but also in the numbers of conferences, open
educational resources, meetups, unconferences, and companies taking up,
as exemplified by the useR! conference
series\footnote{\href{https://www.r-project.org/conferences/}{https://www.r-project.org/conferences/}},
the global growth of the R and R-Ladies user
groups\footnote{\href{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}{https://www.r-consortium.org/blog/2019/09/09/r-community-explorer-r-user-groups}, \href{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}{https://www.r-consortium.org/blog/2019/08/12/r-community-explorer}},
or the foundation and impact of the
R~Consortium\footnote{\href{https://www.r-consortium.org/news/announcements}{https://www.r-consortium.org/news/announcements}, \href{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}{https://www.r-consortium.org/blog/2019/11/14/data-driven-tracking-and-discovery-of-r-consortium-activities}}.
All this cements the role of R as the \emph{lingua~franca} of
statistics, data visualisation, and computational research. Coinciding
with the rise of R was the advent of
\href{https://en.wikipedia.org/wiki/Docker_(software)}{Docker} as a
general tool for distribution and deployment of server applications---in
fact, Docker can be called the \emph{lingua~franca} of describing
computing environments and packaging software. Combining both these
topics, the \emph{Rocker~Project}
(\url{https://www.rocker-project.org/}) provides images with R (see the
next Section for more details). The considerable uptake and continued
evolution of the Rocker~Project has lead to numerous projects extending
or building upon Rocker images, ranging from reproducible
research\textbackslash{}footnote\{``Reproducible'' in the sense of the
\emph{Claerbout/Donoho/Peng} terminology
\citep{barba_terminologies_2018}.\} to production deployments. This
article presents this \emph{Rockerverse} of projects across all
development stages: early demonstrations, working prototypes, and mature
products. We also introduce related activities connecting the R language
and environment with other containerisation solutions. The main
contributions is a coherent picture of the current lay of the land of
using containers in, with, and for R.

The article continues with a brief introduction of containerization
basics and the Rocker~Project, followed by descriptions of projects
developing images with R installations. Then, we present applications,
starting with the R packages specificly for interacting with Docker,
second-level packages using containers mediately or only for specific
features, up to complex use cases leveraging containers. We conclude
with a reflection on the landscape of packages and applications and
point out future directions of development.

@@

Docker, an application and service provide by the eponymous company, has
in just a few short years risen to prominence for development, testing,
deployment and distribution of computer software
\citep[cf.][]{datadog_8_2018,munoz_history_2019}. While there are
related approaches such as
LXC\footnote{\href{https://en.wikipedia.org/wiki/LXC}{https://en.wikipedia.org/wiki/LXC}}
or Singularity \citep{kurtzer_singularity_2017}, Docker has become
synomymous with ``containerization''---the method of taking software
artefacts and bundling them in such a way that use becomes standardized
and portable across operationg systems. In doing so, Docker had
recognised and validated the importance of one very important thread
that had been emerging, namely virtualization. By allowing (one or
possibly) multiple applications or services to run concurrently on one
host machine without any fear of interference between them, an important
scalability opportunity is being provided. But Docker improved this
compartmentalization by accessing the host system---generally
Linux---through a much thinner and smaller shim than a full operating
system emulation or virtualization. This containerization is also called
operating-system-level virtualization
\citep{wikipedia_contributors_os-level_2020}. Typically a container runs
one process, whereas virtualization may run whole operating systems at a
larger footprint. This makes for more efficient use of system resources
\citep{felter_updated_2015} and allowed another order of magnitude in
terms of scalability of deployment \citep[cf.][]{datadog_8_2018}. While
Docker makes use of Linux kernel features, its importance was large
enough so that some required aspects of running Docker have been added
to other operating systems to support Docker there more efficiently too
\citep{microsoft_linux_2019}. The success even lead to standardisation
and industry collaboration \citep{oci_open_2019}.

@@

For statistical computing and analysis centered around R, the
Rocker~Project has provided a variety of Docker containers since its
start in 2014 \citep{RJ-2017-065}. The Rocker~Project provides several
lines of containers spanning to from building blocks with
\texttt{R-release} or \texttt{R-devel}, via containers with
\href{https://rstudio.com/products/rstudio/}{RStudio~Server} and
\href{https://rstudio.com/products/shiny/shiny-server/}{Shiny~Server},
to domain-specific containers such as
\href{https://github.com/rocker-org/geospatial}{\texttt{rocker/geospatial}}
\citep{rocker_geospatial_2019}. These containers form \emph{image
stacks}, building on top of each other for better maintainability
(i.e.~smaller \texttt{Dockerfile}s), composability, and to reduce build
time. Also of note is a series of ``versioned'' containers which match
the R release they contain with the \emph{then-current} set of packages
via the MRAN Snapshot views of CRAN \citep{microsoft_cran_2019}. The
Rocker~Project's impact and importance was acknowledged by the Chan
Zuckerberg Initiative's \emph{Essential Open Source Software for
Science}, who provide findung for the projects's sustainable
maintenance, community growth, and targeting new hardware platforms
including GPUs \citep{chan_zuckerberg_initiative_maintaining_2019}.

@@

As outlined above, R is a widely-used language with a large community.
The large number of extension packages provides access to an unrivaled
variety of established and upcoming features. Nevertheless, special use
cases and experimental projects in academia and industry exist to test
new approaches or provide features different to what \emph{GNU-R}, or
Base R, provides. GNU-R is the R distribution maintained by the
\href{https://www.r-project.org/contributors.html}{R Core Team} and
available via \href{https://cran.r-project.org/}{CRAN}. While the images
presented in this section are far from being as vetted, stable, and
widely used as any of the Rocker images, they demonstrate an important
advantage of containerisation technology, namely the abiliy to
transparently build portable stacks of open source software and make
them easily accessible to others. All these distributions of R are
published under GPL licenses. Since all of the R distributions claim
better performance as a core motivation, a comparision based on Docker
images, potentially leveraging the
\href{https://docs.docker.com/config/containers/resource_constraints/}{resource
restriction mechanisms} of Docker to level the playing field, seems
useful future work.

\textbf{Microsoft R Open} (MRO) is an R distribution formerly known as
Revolution R Open (RRO) before Revolution Analytics was acquired by
Microsoft. MRO is compatible with main R and its packages and it
``{[}..{]} includes additional capabilities for improved performance,
reproducibility, and platform support.'' \citep{microsoft_mro} Most
notably these capabilities are the MRAN repository, which is enabled by
default, and the (optional) integration with
\href{https://software.intel.com/en-us/mkl}{Intel® Math Kernel Library}
(MKL) for multi-threading in linear algebra operations
\citep{microsoft_multithread}. MRO does not provide official Docker
images, but a set of community-maintained \texttt{Dockerfile}s and
Docker images are provided by the poject \texttt{mro-docker}
(\url{https://github.com/nuest/mro-docker}), e.g.~the image
\href{https://hub.docker.com/repository/docker/nuest/mro/}{\texttt{nuest/mro}}.
The images are inspired by the Rocker images and can be used much in the
same fashion, effectively a drop-in replacement allowing users to
quickly evaluate if the benefits of MRO + Intel® MKL apply to their use
case. Version-tagged images are provided for the latest bugfix release
of recent R versions. Extended license information about MKL is printed
at every startup.

\textbf{Renjin} is an interpreter for the R language running on top of
the \href{https://en.wikipedia.org/wiki/Java_virtual_machine}{Java
Virtual Machine} (JVM) providing full two-way access between Java and R
code \citep{wikipedia_renjin_2018}. It was developed to combine the
benefits of R, such scripting and extension packages, with the JVM's
advantages in the areas of security, cross-platform availability, and
established position in enterprise settings. R extension packages need
to be specially compiled and are distributed via the Java package
manager \href{https://en.wikipedia.org/wiki/Apache_Maven}{Apache Maven},
cf.~\url{http://packages.renjin.org/packages} for available packages.
Packages are loaded on demand, i.e., at the first call to
\texttt{library()}. Not all R packages, especially one linking to binary
libraries, are available, e.g., \texttt{rgdal}
\footnote{\href{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}{http://packages.renjin.org/package/org.renjin.cran/rgdal/1.4-4/build/1}}.
There are no offical Docker images for Renjin, but community-maintained
images for selected releases only are available under
\texttt{nuest/renjin} on Docker~Hub and GitHub at
\url{https://hub.docker.com/r/nuest/renjin} and
\url{https://github.com/nuest/renjin-docker} respectively. These images
expose the command line interface of Renjin in a similar fashion as
Rocker images and allow an easy evaluation of Renjin's suitability, but
are not intended for production use.

\textbf{pqR} (\url{http://www.pqr-project.org/}) is a ``a pretty quick
version of R''. \emph{pqR} attempts to improve R on some opinionated
issues in the R language and is the basis for experimental features,
e.g., automatic
diffentiation\footnote{\href{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}{https://riotworkshop.github.io/abstracts/riot-2019-pqr.txt}}.
The source code development
\href{https://github.com/radfordneal/pqR/}{on GitHub} is a one man
project and it does not provide any Docker images. But especially
disruptive approaches may contribute to the development of the R
ecosystem, so the \texttt{nuest/pqr} image was independently created,
see Docker Hub at \url{https://hub.docker.com/r/nuest/pqr/} and GitHub
at \url{https://github.com/nuest/pqr-docker}.

\textbf{FastR} (\url{https://github.com/oracle/fastr}) is ``A
high-performance implementation of the R programming language, built on
GraalVM'' \citep{oracle_labs_oraclefastr_2020}. It is developed by
Oracle, connects R to the GraalVM ecosystem
\citep{wikipedia_graalvm_2019}, and also claims superior performance but
also targets full compatibility with base R
\citep{oracle_labs_oraclefastr_2020}. There are no offical Docker images
provided, but indepedently maintained experimental images and
\texttt{Dockerfile}s are provided by
\href{https://github.com/nuest/fastr-docker}{nuest/fastr-docker},
e.g.~\href{https://cloud.docker.com/repository/docker/nuest/fastr/}{\texttt{nuest/fastr}
on Docker Hub}.

@@

\emph{Bioconductor} (\url{https://bioconductor.org/}) is an open source,
open development project for the analysis and comprehension of genomic
data \citep{gentleman_bioconductor_2004}. The project consists of 1741 R
software packages as of August 15th 2019, as well as packages containing
annotation or experiment data. \emph{Bioconductor} has a semiannual
release cycle, each release is associated with a particular version of
R. Docker images allow availability of current and past versions of
\emph{Bioconductor} for convenience and reproducibility.
\emph{Bioconductor}'s `base' docker images are built on top of
\texttt{rocker/r-ver} and \texttt{rocker/rstudio} images.
\emph{Bioconductor} installs packages based on the R version, and
therefore uses \texttt{rocker/rstudio} and \texttt{rocker/r-ver} version
tagging. \emph{Bioconductor} selects the desired version of R from
Rocker, adds the BiocManager CRAN package for installing appropriate
versions of \emph{Bioconductor} packages, and creates a
\emph{Bioconductor} docker image with an informative tag
(R\_version\_Bioc\_version). The images are summarized on the
\emph{Bioconductor} web site (see
\url{https://bioconductor.org/help/docker/}), created with
\texttt{Dockerfile}s maintained on GitHub
(\url{https://github.com/Bioconductor/bioconductor_docker}), and
\href{https://hub.docker.com/u/bioconductor}{distributed through
DockerHub}. Past and current combinations of R and \emph{Bioconductor}
are therefore accessible via a specific image tag. A strength of this
approach is that the responsibility for complex software configuration
(including customized development) is shifted from the user to the
experienced \emph{Bioconductor} core team. Images may be valuable in
teaching, where participants pull pre-built images to avoid complicated
configuration of their own computing environemnts.

@@

@@

Docker containers on the Windows operating system were originally quite
cumbersome to use. They required an extra tool,
\href{https://docs.docker.com/toolbox/}{Docker Toolbox}, which by now
only exists as a legacy solution for older Windows systems. Docker
toolbox leverages \texttt{docker-machine} (see also
Section~\nameref{interfaces} for an R package interfacing with
\texttt{docker-machine}) to handle the process of creating a local
virtual host which could host Docker Engine, while exposing the regular
Docker CLI. The Docker CLI commands are forwarded to the virtual host
transparently for the user.

For current Windows Server (2016 and later) and Windows Desktop
(\href{https://www.docker.com/products/docker-desktop}{Docker Desktop}
requires Windows 10) versions, Docker is supported
natively\footnote{\href{https://www.docker.com/products/windows-containers}{https://www.docker.com/products/windows-containers}}
and different base images are offered by
Microsoft\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/about/index}}.
The Docker CLI can be used in just the same way as on other operating
systems but not every base image is supported on every Windows
host\footnote{\href{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}{https://docs.microsoft.com/en-us/virtualization/windowscontainers/deploy-containers/version-compatibility}}.
On Docker Desktop for Windows, the user can run both Linux-based and
Windows-based containers, but only one of the Docker daemons can be used
at a
time\footnote{See "Switch between Windows and Linux containers" on \href{https://docs.docker.com/docker-for-windows/}{https://docs.docker.com/docker-for-windows/}.}.

\textbf{rocker-win} (\url{https://github.com/nuest/rocker-win}) is a
proof of concept for running R in Windows-based containers
\citep{nust_rocker-win_2019}. It provides selected R versions using
three different base images using \texttt{microsoft/windowsservercore}
and \texttt{mcr.microsoft.com/windows/servercore}. These base images
match different Windows versions: Windows Server 2019, Windows Server
2016, and Windows Server, version 1803. The images are built using two
CI services,
\href{https://docs.travis-ci.com/user/reference/windows/}{Travis CI} and
\href{https://www.appveyor.com/docs/windows-images-software/}{Appveyor},
which provide the different Windows versions and therefore support
different base images, and published automatically
\href{https://hub.docker.com/r/nuest/rocker-win}{on Docker~Hub}. The
images are built from manually maintained seperate \texttt{Dockerfile}s,
are tagged with both Windows-variant and R version, e.g.,
\texttt{nuest/rocker-win:ltsc2019-3.6.2} or
\texttt{nuest/rocker-win:1803-latest}, and run the \texttt{R.exe}
process by default. All these images can be run on Docker for Desktop on
Windows 10. New R versions are only added on demand. The explorative
rocker-win project demonstrates the variety and pontentially confusing
complexity of Windows on Docker, e.g., because of the different base
images, but also provides a starting point for any Windows-constrained
user to leverage containerisation, including examples for apps using
\textbackslash{}pkg\{plumber, \pkg{Shiny}, or R packages with system
dependencies such as \pkg{sf}. The latter is quite comfortable and fast
actually, compared to installation from source on Linux, as R ships
pre-compiled binaries for Windows. However, compared to the various
Linux images, the footprint of Windows images is quite large.

For Windows, just as originally for Linux, the cloud use cases drive the
development and Docker and Microsoft collaborate
closely\footnote{\href{https://www.docker.com/partners/microsoft}{https://www.docker.com/partners/microsoft}}.
In most cases, individual developers or researchers worrying about
reproducibility, will prefer the more widely and host independent
Linux-based containers. Naturally, a Windows license is required for the
host machine, and the licensing impacts the potnetial to build images in
clouds or redistribute exported images as archive files. But the
developments in the \texttt{rocker-win} prototype show that containers
can also be used for R workflows depending on Windows-only tools, for
leveraging an existing Windows Server-based infrastructure where
policies can otherwise not be met, or for streamlining interactions with
system operating staff that runs Windows-based containers themselves.

@@

\textbf{\pkg{renv}} (\url{https://rstudio.github.io/renv/}) with Docker:
\url{https://rstudio.github.io/renv/articles/docker.html}

\emph{version pinning} with packages \pkg{remotes} and \pkg{versions}
for R, and with system package managers for distros

@@

  \begin{itemize}
  \tightlist
  \item
    For this project, we use travis-ci to build rocker-derived
    containers, test them, and only push them to Docker~Hub (from
    travis-ci.org) if the integration tests succeed.
  \end{itemize}
\item
  Google Cloud Build

@@

\textbf{RCloud} (\url{https://rcloud.social}) is a cloud-based platform
for data analysis, visualisation and collaboration using R. It provides
a \texttt{rocker/drd} base image for easy evaluation of the
platform\footnote{\href{https://github.com/att/rcloud/tree/master/docker}{https://github.com/att/rcloud/tree/master/docker}}.

\hypertarget{using-docker-in-a-commercial-data-science-platform}{%
\subsection{Using Docker in a commercial data science
platform}\label{using-docker-in-a-commercial-data-science-platform}}

@@

\hypertarget{processing}{%
\subsection{Processing}\label{processing}}
